---
title: "wood_dynamism"
author: 'Noël Clark'
date: "2026-06-22"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      out.width = "100%")
```

```{r}
source("setup.R")
```


```{r load libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(plotly)
library(patchwork)
library(GGally)
library(lme4)
library(DHARMa)
library(lmerTest)
library(performance)
library(MASS)
library(see)
library(AICcmodavg)
library(ggeffects)
library(sjPlot)
library(emmeans)
```

```{r kelley load data, message = FALSE}
setwd("~/Library/CloudStorage/OneDrive-TheUniversityofMontana/Wood+/Wood+") # Kelley's working directory

dynamism_data <- read_csv("site_metrics_07022026.csv") # Kelley's path
data_byyr <- read_csv("SelectMetricsByYear_07022026.csv") # Kelley's path
all_data_byyr <- read.csv("ALL_metrics_07022026.csv") # Kelley's path
```

```{r noel load data, message = FALSE}
## change directory if needed ##
data_dir <- "C:/Users/nc274936/OneDrive - The University of Montana/wood_dynamism/data"

dynamism_data <- read_csv(file.path(data_dir, "Metrics/site_metrics_07022026.csv"))
data_byyr <- read_csv(file.path(data_dir, "/Metrics/SelectMetricsByYear_07022026.csv"))
all_data_byyr <- read.csv(file.path(data_dir, "/Metrics/ALL_metrics_07022026.csv"))
```

```{r clean loaded data}
## this should work after you run the right version of data loading above
dynamism_data <- dynamism_data[1:13,] %>% # for some reason we have duplicate rows in the csv
  bind_rows( # also want to add back the mod rock_WA site
    dynamism_data %>% slice(6) %>% 
      mutate(across(
        c(site_reach, site),
        ~ str_replace_all(.x, "_1", "_2")
        ))
    ) 

data_byyr <- data_byyr %>% 
  rename(reach = site,
         site = river) %>% 
  mutate(site_reach = paste0(site,reach)) %>% 
  mutate(ratio_accum_to_ch_area = accum_area/channel_area,
         piece_per_100m = (wood_pieces/channel_length)*100,
         site = factor(site),
         reach = factor(reach))

all_data_byyr <- all_data_byyr %>% 
  rename(reach = site,
         site = river) %>% 
  mutate(site_reach = paste0(site,reach))
```

# Simple Plots
 
plot settings
```{r plot settings}
wood_colors = c("LW_Channel" = "#99C5B5", "no_LW_Channel" = "#BEA57D")
```

# visualizing distribution of values 

support for the response values is positive & continuous
none of the response values are normally distributed - justifies generalized linear model with gamma family and log link

```{r simple value density plots}
data_byyr %>%
  filter(site_reach != "Rock_WA_2no_LW_Channel") %>% 
  ggplot(aes(x=sinuosity)) +
  geom_density() 

data_byyr %>% 
  filter(site_reach != "Rock_WA_2no_LW_Channel") %>% 
  ggplot(aes(x=complexity)) +
  geom_density()

data_byyr %>%
  filter(site_reach != "Rock_WA_2no_LW_Channel") %>% 
  ggplot(aes(x=ratio_accum_to_ch_area)) +
  geom_density() 

data_byyr %>% 
  filter(site_reach != "Rock_WA_2no_LW_Channel") %>% 
  ggplot(aes(x=piece_per_100m)) +
  geom_density() 
```

## wood vs. site, reach, & time
```{r wood v time plots}
# how does standardized wood (the ratio of wood accumulation to channel area) change over time?
data_byyr %>%
  ggplot(aes(x=year, y=ratio_accum_to_ch_area, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach))

# how does the difference between standardized wood change within a site (between reaches) over time? (diff = LW - noLW)
data_byyr %>%
  group_by(site, year) %>% 
  arrange_all(by_group = "reach") %>% 
  mutate(accum_ratio_diff = lag(ratio_accum_to_ch_area, 1) - ratio_accum_to_ch_area) %>% 
  dplyr::select(site, year, accum_ratio_diff) %>% 
  drop_na(accum_ratio_diff) %>% 
  ggplot(aes(x=year, y=accum_ratio_diff, color = site)) +
  geom_point() +
  geom_line(aes(group = site))


# how does standardized wood pieces change over time?
data_byyr %>% 
  ggplot(aes(x=year, y=piece_per_100m, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach))

# how does the difference between standardized number of wood pieces change within a site (between reaches) over time? (diff = LW - noLW)
data_byyr %>%
  group_by(site, year) %>% 
  arrange_all(by_group = "reach") %>% 
  mutate(piece_100m_diff = lag(piece_per_100m, 1) - piece_per_100m) %>% 
  dplyr::select(site, year, piece_100m_diff) %>% 
  drop_na(piece_100m_diff) %>% 
  ggplot(aes(x=year, y=piece_100m_diff, color = site)) +
  geom_point() +
  geom_line(aes(group = site))
```
```{r simple wood site plots}
##### wood values by site by year #####
data_byyr %>% # Cleaned up this plot a bit
  ggplot(aes(x = year, y = ratio_accum_to_ch_area, color = reach)) +
  geom_point(size = 2) +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors,
          name = "Reach Type",
          labels = c("Large Wood", "Control")) +
  labs(y = "Wood accumulation : Channel area ratio") + 
  theme(legend.position = "right",
        panel.grid = element_blank())+
    theme_bw() 

data_byyr %>% 
  ggplot(aes(x = year, y = piece_per_100m, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

##### mean wood values by site #####
dynamism_data %>% 
  ggplot(aes(x = site, y = avg_wood_pieces_per100m, fill = reach)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = wood_colors) +
  theme(legend.position = "none")

dynamism_data %>% 
  ggplot(aes(x = site, y = mean_wood_accum_per_area, fill = reach)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = wood_colors)

```


## sinuosity, complexity, and sdi vs. standardized wood

visualizing relationship between response v predictor (ratio_accum_to_ch_area)
within a site the relationship is linear-ish? 
```{r visualize sin comp and sdi v wood}
data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area, y=sinuosity, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site))

data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area, y=complexity, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site))

# data_byyr %>% 
#   ggplot(aes(x=ratio_accum_to_ch_area, y=sdi, color = site, shape = reach)) +
#   geom_point() +
#   geom_line(aes(group=site))


## wood piece_per_100m
data_byyr %>% 
  ggplot(aes(x=piece_per_100m, y=sinuosity, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site))

data_byyr %>% 
  ggplot(aes(x=piece_per_100m, y=complexity, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site))

# data_byyr %>% 
#   ggplot(aes(x=piece_per_100m, y=sdi, color = site, shape = reach)) +
#   geom_point() +
#   geom_line(aes(group=site))
```

channel values by site
```{r simple channel site plots}
# channel characteristics v time by site
data_byyr %>% 
  ggplot(aes(x = year, y = channel_length, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

data_byyr %>% 
  ggplot(aes(x = year, y = channel_area, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) + 
  scale_color_manual(values = wood_colors)
```

response values by site
```{r simple response values by site}
# response metrics v time by site
data_byyr %>% 
  ggplot(aes(x = year, y = sinuosity, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

data_byyr %>% 
  ggplot(aes(x = year, y = complexity, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

# data_byyr %>% 
#   ggplot(aes(x = year, y = sdi, color = reach)) +
#   geom_point() +
#   geom_line(aes(group = site_reach)) +
#   facet_wrap(~site) +
#   scale_color_manual(values = wood_colors)

# all_data_byyr %>% 
#   ggplot(aes(x = year_to, y = net_change, color = reach)) +
#   geom_point() +
#   geom_line(aes(group = site_reach)) +
#   facet_wrap(~site) +
#   scale_color_manual(values = wood_colors)
# 
# all_data_byyr %>% 
#   ggplot(aes(x = year_to, y = erosion_area, color = reach)) +
#   geom_point() +
#   geom_line(aes(group = site_reach)) +
#   facet_wrap(~site) +
#   scale_color_manual(values = wood_colors)
# 
# all_data_byyr %>% 
#   ggplot(aes(x = year_to, y = deposition_area, color = reach)) +
#   geom_point() +
#   geom_line(aes(group = site_reach)) +
#   facet_wrap(~site) +
#   scale_color_manual(values = wood_colors)
```

dynamism values by site
```{r simple dynamism plots}

library(rcartocolor)
rcartocolor::display_carto_all()

##### Sinuosity vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = sinuosity_dynamism, color = site)) +
  geom_point(aes(shape = reach), size = 3) +
    scale_color_carto_d(palette = "Earth") +
    scale_shape_manual(
      values = c(16, 17),
      labels = c("Large Wood", "Control"),
      name = "Reach Type") +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", 
       y = "Sinuosity Dynamism",
       color = "Site",
       shape = "Reach Type")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = sinuosity_dynamism, color = site)) +
  geom_point(aes(shape = reach), size = 3) +
    scale_color_carto_d(palette = "Earth") +
    scale_shape_manual(
    values = c(16, 17),
    labels = c("Large Wood", "Control"),
    name = "Reach Type") +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  theme(legend.position = "right") +
  labs(x = "Avg. Wood Acculumations per Area", 
       y = "",
       color = "Site",
       shape = "Reach Type")

(p1 + p2) +
  plot_annotation(title = "Sinuosity Dynamism vs. Avg Wood Metrics")
##### end 

##### Complexity vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = complexity_dynamism, color = site)) +
  geom_point(aes(shape = reach), size = 3) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "Complexity Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = complexity_dynamism, color = site)) +
  geom_point(aes(shape = reach),size = 3) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "Complexity Dynamism vs. Avg Wood Metrics")
##### end

##### SDI vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = sdi_dynamism, color = site)) +
  geom_point(aes(shape = reach), size = 3) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "SDI Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = sdi_dynamism, color = site)) +
  geom_point(aes(shape = reach), size = 3) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "SDI Dynamism vs. Avg Wood Metrics")
##### end

##### Net change vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = net_change_dynamism, color = site)) +
  geom_point(aes(shape = reach), size = 3) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "Net Change Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = net_change_dynamism, color = site)) +
  geom_point(aes(shape = reach), size = 3) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "Net Change Dynamism vs. Avg Wood Metrics")
##### end
```
```{r simple dynamism plots PART 2 MEGA PANEL}
##### Sinuosity vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = sinuosity_dynamism, color = site)) +
  geom_point(aes(shape = reach), size = 3) +
    scale_color_carto_d(palette = "Earth") +
    scale_shape_manual(
      values = c(16, 17),
      labels = c("Large Wood", "Control"),
      name = "Reach Type") +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "", 
       y = "Sinuosity",
       color = "Site",
       shape = "Reach Type")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = sinuosity_dynamism, color = site)) +
  geom_point(aes(shape = reach), size = 3) +
    scale_color_carto_d(palette = "Earth") +
    scale_shape_manual(
    values = c(16, 17),
    labels = c("Large Wood", "Control"),
    name = "Reach Type") +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = "", 
       y = "",
       color = "Site",
       shape = "Reach Type")

sinuosity <- (p1 + p2) +
  plot_annotation(title = "Sinuosity Dynamism vs. Avg Wood Metrics")
##### end 

##### Complexity vs both mean wood metrics #####
p3 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = complexity_dynamism, color = site)) +
  geom_point(aes(shape = reach), size = 3) +
    scale_color_carto_d(palette = "Earth") +
    scale_shape_manual(
      values = c(16, 17),
      labels = c("Large Wood", "Control"),
      name = "Reach Type") +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "", 
       y = "Complexity",
       color = "Site",
       shape = "Reach Type")

p4 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = complexity_dynamism, color = site)) +
  geom_point(aes(shape = reach), size = 3) +
    scale_color_carto_d(palette = "Earth") +
    scale_shape_manual(
    values = c(16, 17),
    labels = c("Large Wood", "Control"),
    name = "Reach Type") +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  theme(legend.position = "right") +
  labs(x = "", 
       y = "",
       color = "Site",
       shape = "Reach Type")

complexity <- (p3 + p4) +
  plot_annotation(title = "Complexity Dynamism vs. Avg Wood Metrics")
##### end

##### SDI vs both mean wood metrics #####
p5 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = sdi_dynamism, color = site)) +
  geom_point(aes(shape = reach), size = 3) +
    scale_color_carto_d(palette = "Earth") +
    scale_shape_manual(
      values = c(16, 17),
      labels = c("Large Wood", "Control"),
      name = "Reach Type") +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "", 
       y = "SDI",
       color = "Site",
       shape = "Reach Type")

p6 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = sdi_dynamism, color = site)) +
  geom_point(aes(shape = reach), size = 3) +
    scale_color_carto_d(palette = "Earth") +
    scale_shape_manual(
    values = c(16, 17),
    labels = c("Large Wood", "Control"),
    name = "Reach Type") +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = "", 
       y = "",
       color = "Site",
       shape = "Reach Type")

sdi <- (p5 + p6) +
  plot_annotation(title = "SDI Dynamism vs. Avg Wood Metrics")
##### end

##### Net change vs both mean wood metrics #####
p7 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = net_change_dynamism, color = site)) +
  geom_point(aes(shape = reach), size = 3) +
    scale_color_carto_d(palette = "Earth") +
    scale_shape_manual(
      values = c(16, 17),
      labels = c("Large Wood", "Control"),
      name = "Reach Type") +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", 
       y = "Net Change",
       color = "Site",
       shape = "Reach Type")

p8 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = net_change_dynamism, color = site)) +
  geom_point(aes(shape = reach), size = 3) +
    scale_color_carto_d(palette = "Earth") +
    scale_shape_manual(
    values = c(16, 17),
    labels = c("Large Wood", "Control"),
    name = "Reach Type") +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Acculumations per Area", 
       y = "",
       color = "Site",
       shape = "Reach Type")

netdyno <- (p7 + p8) +
  plot_annotation(title = "Net Change vs. Avg Wood Metrics")


# Patching together
(p1 + p2 +
 p3 + p4 +
 p5 + p6 +
 p7 + p8) +
  plot_layout(ncol = 2) +
  plot_annotation(title = "Dynamism vs. Avg Wood Metrics")
##### end

```

```{r simple dynamism plots pt. 2}
library(grid)
library(rcartocolor)
library(dplyr)
library(patchwork)

## CLEANING UP
# Adjusting so all 3 metrics can be plotted on same graph
dynamism_cubed <- dynamism_data %>%
  pivot_longer(c(complexity_dynamism,sinuosity_dynamism, sdi_dynamism, net_change_dynamism),
               names_to = "Metric",
               values_to = "Value") %>%
  group_by(Metric) %>%
  mutate(Value_scaled = (scale(Value))) # calculated a separate z-score for each dynamism metric
  # will show dynamism values relative to each other

# Essentially, does wood accumulation change dynamism relative to it's normal range?

# Looking at colorblind friendly color schemes
#rcartocolor::display_carto_all()
#carto_pal(7, "Earth")

# Renaming labels
dynamism_cubed <- dynamism_cubed %>%
  mutate(
    Metric = recode(
      Metric,
      complexity_dynamism = "Complexity",
      sinuosity_dynamism = "Sinuosity",
      sdi_dynamism = "SDI",
      net_change_dynamism = "Net Channel Change"
    ),
    reach = recode(
      reach,
      LW_Channel = "Large Wood",
      no_LW_Channel = "Control"
    ))
    
## PLOTS

# Mean wood pieces (pooled)
ggplot(dynamism_cubed, aes(avg_wood_pieces_per100m, Value_scaled, color = Metric)) +
  geom_line(size = 1) +
  scale_color_manual(values = c(
  "Complexity" = "#B5B991",
  "Sinuosity" = "#70A494",
  "SDI" = "#DE8A5A",
  "Net Channel Change" = "#E15759"
  )) +
  labs(x = "Mean wood pieces (per 100m)",
    y = "Standardized dynamism (z-score)") +
  theme_classic()

# Mean wood accumulation (pooled)
ggplot(dynamism_cubed, aes(mean_wood_accum_per_area, Value_scaled, color = Metric)) +
  geom_line(size = 1)+
  scale_color_manual(values = c(
  "Complexity" = "#B5B991",
  "Sinuosity" = "#70A494",
  "SDI" = "#DE8A5A",
  "Net Channel Change" = "#E15759"
  )) +
  labs(x = "Mean wood accumulation (per area)",
    y = "Standardized dynamism (z-score)") +
  theme_classic()

## COMBO PLOT
# Mean wood peices across site and reach
meanwoodpieces <- ggplot(
  dynamism_cubed,
  aes(
    x = avg_wood_pieces_per100m,
    y = Value_scaled,
    color = Metric
  )
) +
  scale_color_manual(values = c(
  "Complexity" = "#B5B991",
  "Sinuosity" = "#70A494",
  "SDI" = "#DE8A5A",
  "Net Channel Change" = "#E15759"
  ))+
 geom_point(
    aes(shape = site),
    size = 3)+
  scale_shape_manual(values = c(16, 17, 15, 18, 3, 8, 4)) +
  geom_smooth(aes(group = Metric),
              method = "lm",
              se = FALSE,
              linewidth = 1.2) +
  facet_wrap(~reach) +
    facet_wrap(~reach, scales = "free_x") + # delete scale = "free_x" if you want each reach on the same scale
    labs(
    x = "Mean wood pieces (per 100m)",
    y = "Standardized dynamism (z-score)",
    color = "Dynamism Metric",
    shape = "Site")+
  theme_bw()


# Mean wood accumulation across site and reach
meanwoodaccum <- ggplot(
  dynamism_cubed,
  aes(
    x = mean_wood_accum_per_area,
    y = Value_scaled,
    color = Metric
  )
) +
  scale_color_manual(values = c(
  "Complexity" = "#B5B991",
  "Sinuosity" = "#70A494",
  "SDI" = "#DE8A5A",
  "Net Channel Change" = "#E15759"
  ))+
 geom_point(
    aes(shape = site),
    size = 3)+
  scale_shape_manual(values = c(16, 17, 15, 18, 3, 8, 4)) +
  geom_smooth(aes(group = Metric),
              method = "lm",
              se = FALSE,
              linewidth = 1.2) +
    facet_wrap(~reach, scales = "free_x") + # delete scale = "free_x" if you want on the same scale
    labs(
    x = "Mean wood accumulation (per area)",
    y = "Standardized dynamism (z-score)",
    color = "Dynamism Metric",
    shape = "Site")+
  theme_bw()+
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )


## FINAL PLOT
(meanwoodpieces + meanwoodaccum) +
  plot_layout(guides = "collect") +
  plot_annotation(title = "Dynamism vs. Average Wood Metrics") &
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    legend.key.size = unit(0.4, "cm"),
    legend.spacing.x = unit(0.2, "cm"),
    legend.text = element_text(size = 9),
    legend.box.spacing = unit(0.1, "cm"),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  ) &
  guides(
    color = guide_legend(
      title = "Dynamism Metric",
      ncol = 1
    ),
    shape = guide_legend(
      title = "Site",
      ncol = 1,
      override.aes = list(size = 3)
    )
  )



## NOW, ON THE SAME X SCALE

meanwoodpieces_samex <- ggplot(
  dynamism_cubed,
  aes(
    x = avg_wood_pieces_per100m,
    y = Value_scaled,
    color = Metric
  )
) +
  scale_color_manual(values = c(
  "Complexity" = "#B5B991",
  "Sinuosity" = "#70A494",
  "SDI" = "#DE8A5A",
  "Net Channel Change" = "#E15759"
  ))+
 geom_point(
    aes(shape = site),
    size = 3)+
  scale_shape_manual(values = c(16, 17, 15, 18, 3, 8, 4)) +
  geom_smooth(aes(group = Metric),
              method = "lm",
              se = FALSE,
              linewidth = 1.2) +
  facet_wrap(~reach) +
    facet_wrap(~reach) + # delete scale = "free_x" if you want each reach on the same scale
    labs(
    x = "Mean wood pieces (per 100m)",
    y = "Standardized dynamism (z-score)",
    color = "Dynamism Metric",
    shape = "Site")+
  theme_bw()


meanwoodaccum_samex <- ggplot(
  dynamism_cubed,
  aes(
    x = mean_wood_accum_per_area,
    y = Value_scaled,
    color = Metric
  )
) +
  scale_color_manual(values = c(
  "Complexity" = "#B5B991",
  "Sinuosity" = "#70A494",
  "SDI" = "#DE8A5A",
  "Net Channel Change" = "#E15759"
  ))+
 geom_point(
    aes(shape = site),
    size = 3)+
  scale_shape_manual(values = c(16, 17, 15, 18, 3, 8, 4)) +
  scale_x_continuous(
  labels = scales::label_number(accuracy = 0.1) # putting this panel to the same decimal point as mean wood peices
)+
  geom_smooth(aes(group = Metric),
              method = "lm",
              se = FALSE,
              linewidth = 1.2) +
    facet_wrap(~reach) + 
    labs(
    x = "Mean wood accumulation (per area)",
    y = "Standardized dynamism (z-score)",
    color = "Dynamism Metric",
    shape = "Site")+
  theme_bw()+
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

(meanwoodpieces_samex + meanwoodaccum_samex) +
  plot_layout(guides = "collect") +
  plot_annotation(title = "Dynamism vs. Average Wood Metrics") &
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    legend.key.size = unit(0.4, "cm"),
    legend.spacing.x = unit(0.2, "cm"),
    legend.text = element_text(size = 9),
    legend.box.spacing = unit(0.1, "cm"),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  ) &
  guides(
    color = guide_legend(
      title = "Dynamism Metric",
      ncol = 1
    ),
    shape = guide_legend(
      title = "Site",
      ncol = 1,
      override.aes = list(size = 3)
    )
  )

```


# correlation plots and test 

ggpairs uses the pearson correlation test by default
```{r correlation plot test}
data_byyr %>%
  filter(site_reach != "Rock_WA_2no_LW_Channel") %>% 
  dplyr::select(sinuosity, complexity, sdi, channel_length, channel_area, accum_area, ratio_accum_to_ch_area, wood_pieces, piece_per_100m) %>% 
  ggpairs(progress = F)

dynamism_data %>% 
  filter(site_reach != "Rock_WA_2no_LW_Channel") %>% 
  dplyr::select(sinuosity_dynamism, complexity_dynamism, sdi_dynamism, net_change_dynamism, avg_wood_pieces_per100m, mean_wood_accum_per_area) %>% 
  ggpairs(progress = F)
```

# Models

```{r create model df}
mod_data_byyr <- data_byyr %>% 
  mutate(ratio_accum_scaled = scale(ratio_accum_to_ch_area, 
                                    center = TRUE, # centers ratio values around mean ratio which is then set to 0 
                                    scale = TRUE), # scale changes a value to the # of sd from the mean value
         piece_100m_scaled = scale(piece_per_100m, 
                                   center = TRUE, 
                                   scale = TRUE),
         year_c = scale(year, # centers years around a mean (set to 0) value for year, makes for a more interpretable effect
                        center = TRUE, scale = FALSE)) %>% 
  filter(site_reach != "Rock_WA_2no_LW_Channel") %>% 
  mutate(site = str_replace_all(site, "_[:digit:]$", "")) %>% 
  mutate(site = factor(site))
```

## specify models

notes
- "random effects should reflect the design, but complexity should be justified by the amount of information available"
- all of our response variables are not normally distributed so we're using a generalized linear mixed model with either a Gamma family(log link) or a gaussian family(log link). 

- null model: random effect of site only reflects hierarchical design 
  - this will estimate a random intercept of sinuosity for every site (i.e. mean sinuosity per site)

- alt models test effect of scaled ratio of wood accumulation
  - alt model 1 uses a gamma log link
    - m1 <- glmer(sinuosity ~ ratio_accum_scaled + (1|site), mod_data_byyr, family = Gamma(link = "log"))
  - alt model 2 uses a gaussian log link
    - m2 <- glmer(sinuosity ~ ratio_accum_scaled + (1|site), mod_data_byyr, family = gaussian(link = "log"))

- we considered models with more complicated random effects structure that estimated both a random intercept and a random slope within a site; however, on second thought we realized we didn't need them because sites are considered a "random" sample of the global population of sites and we're interested in the effect of wood across this global population of sites, not within sites

```{r specify all models together}
responses <- c("sinuosity", "complexity")

fit_candidates <- function(response, data) {
  list(
    null = glmer(reformulate("(1|site)", response), 
                 data, family = Gamma("log")),
    m1   = glmer(reformulate("ratio_accum_scaled + (1|site)", response), 
                 data, family = Gamma("log")),
    m2   = glmer(reformulate("ratio_accum_scaled + (1|site)", response), 
                 data, family = gaussian("log"))
  )
}

models <- map(set_names(responses), fit_candidates, data = mod_data_byyr)
```

## model selection using AICc
- model selection: AICc is preferred for small sample sizes which is when the number of data points / number of parameters < 40 (n/k < 40). we have 126 observations, which for the null model means we're not in the small sample size world (k=3), but for all other models we are because we have to estimate at least 4 parameters. it's good to use the same criteria across models so we'll use AICc for all. the aictab function in the AICcmodavg does this for us.
```{r model selection}
# m1 is the top model for sinuosity
AICcmodavg::aictab(models$sinuosity)
# m2 is the top model for complexity 
AICcmodavg::aictab(models$complexity)

# look at the top model summaries
## these estimates are on the log scale
summary(models$sinuosity$m1)
summary(models$complexity$m2)

## these are both on the log scale
#tab_model(models$sinuosity$m1) 
#tab_model(models$complexity$m2) 
```

## obtain estimates on original response scale
still working on this section
```{r top model estimates}
emmeans(models$sinuosity$m1, ~ ratio_accum_scaled, type = "response")

```


## plot model predictions
```{r plot sinuosity model predictions}
# sinuosity : top model is m1, then m2, then null

## prediction plots for m1 (top model)
plot(
  ggpredict(
    models$sinuosity$m1, terms = c("ratio_accum_scaled", "site"), type = "random", bias_correction = TRUE
    )
  ) +
  geom_point(
    data = mod_data_byyr,
    aes(x = ratio_accum_scaled, y = sinuosity, color = site),
    alpha = 0.5, inherit.aes = FALSE
  ) +
  labs(title = "m1 predicted & observed values of sinuosity")

## predicition plots for m2
plot(
  ggpredict(
    models$sinuosity$m2, terms = c("ratio_accum_scaled", "site"), type = "random", bias_correction = TRUE
    )
  ) +
  geom_point(
    data = mod_data_byyr,
    aes(x = ratio_accum_scaled, y = sinuosity, color = site),
    alpha = 0.5, inherit.aes = FALSE
  ) +
  labs(title = "m2 predicted & observed values of sinuosity")

## prediction plots for the null model 

null_sin_pred_df <- as.data.frame(ggpredict(
    models$sinuosity$null, terms = c("site"), type = "random", bias_correction = TRUE
    )
) %>% 
  rename(site = x)

ggplot() +
  geom_point( # observed points
    data = mod_data_byyr,
    aes(x = site, y = sinuosity, color = site), alpha = 0.7
  ) +
  geom_boxplot( # observed means and sd
    data = mod_data_byyr,
    aes(x = site, y = sinuosity, fill = site, color = site), alpha = 0.5
  ) +
  geom_point( # predicted means
    data = null_sin_pred_df,
    aes(x = site, y = predicted),
    color = "black"
    ) +
  geom_errorbar( # predicted upper and lower 95% CIs
    data = null_sin_pred_df,
    aes(x = site, ymin = conf.low, ymax = conf.high), 
    color = "black"
    ) +
  labs(title = "null predicted & observed values of sinuosity")
```

```{r plot complexity model predictions}
# complexity : top model is m2, then null, then m1

## prediction plots for m2 (top model)
plot(
  ggpredict(
    models$complexity$m2, terms = c("ratio_accum_scaled", "site"), type = "random", bias_correction = TRUE
    )
  ) +
  geom_point(
    data = mod_data_byyr,
    aes(x = ratio_accum_scaled, y = complexity, color = site),
    alpha = 0.5, inherit.aes = FALSE
  ) +
  labs(title = "m2 predicted & observed values of complexity")


## prediction plots for the null model 

null_comp_pred_df <- as.data.frame(ggpredict(
    models$complexity$null, terms = c("site"), type = "random", bias_correction = TRUE
    )
) %>% 
  rename(site = x)

ggplot() +
  geom_point( # observed points
    data = mod_data_byyr,
    aes(x = site, y = complexity, color = site), alpha = 0.7
  ) +
  geom_boxplot( # observed means and sd
    data = mod_data_byyr,
    aes(x = site, y = complexity, fill = site, color = site), alpha = 0.5
  ) +
  geom_point( # predicted means
    data = null_comp_pred_df,
    aes(x = site, y = predicted),
    color = "black"
    ) +
  geom_errorbar( # predicted upper and lower 95% CIs
    data = null_comp_pred_df,
    aes(x = site, ymin = conf.low, ymax = conf.high), 
    color = "black"
    ) +
  labs(title = "null predicted & observed values of complexity") 
  
  
## complexity m1 predictions
plot(
  ggpredict(
    models$complexity$m1, terms = c("ratio_accum_scaled", "site"), type = "random", bias_correction = TRUE
    )
  ) +
  geom_point(
    data = mod_data_byyr,
    aes(x = ratio_accum_scaled, y = complexity, color = site),
    alpha = 0.5, inherit.aes = FALSE
  ) +
  labs(title = "m1 predicted & observed values of complexity") 

```

## model diagnostics with DHARMa & performance packages
```{r DHARMa checks for candidate models}
# simulate residuals for sinuosity
sim_resids_sin <- purrr::map(models$sinuosity, 
                         ~DHARMa::simulateResiduals(fittedModel = .x), 
                         .progress = FALSE)

# Plot simulated residuals: (QQ-plot for distribution, Residual vs. Predicted for variance)
purrr::map(sim_resids_sin, 
           ~plot(.x, quantreg = FALSE),
           .progress = FALSE)

# simulate residuals for complexity 
sim_resids_comp <- purrr::map(models$complexity, 
                         ~DHARMa::simulateResiduals(fittedModel = .x), 
                         .progress = FALSE)

# Plot simulated residuals: (QQ-plot for distribution, Residual vs. Predicted for variance)
purrr::map(sim_resids_comp, 
           ~plot(.x, quantreg = FALSE),
           .progress = FALSE)
```

```{r performance checks for candidate models}
# plots various model assumptions (normality of residuals, normality of random effects, linear relationship, homogeneity of variance, multicollinearity).
# some of the models have flat-ish reference lines for homogeneity of variance
purrr::map(models$sinuosity, ~performance::check_model(.x))
purrr::map(models$complexity, ~performance::check_model(.x))
```
## residual plots
```{r residual plots for sinuosity}
sim_resids <- sim_resids_sin

# recalc residuals aggregated by site
recal_sim_resids <- purrr::map(sim_resids, 
                               ~DHARMa::recalculateResiduals(.x, group = ~site),
                               .progress = FALSE)

# plotting QQ plot and predicted vs residuals aggregated by site
# makes for a low sample size of residuals 
#purrr::map(recal_sim_resids, ~plot(.x, quantreg = FALSE))

# plotting just residuals

resid_df <- map2_dfr(
  sim_resids,
  paste0(seq_along(sim_resids)),
  ~ tibble(
    model = .y,
    index = seq_along(residuals(.x)),
    residual = residuals(.x)
  )
) %>% 
  mutate(model = case_when(
    model == "1" ~ "null_sin",
    model == "2" ~ "m1_sin",
    model == "3" ~ "m2_sin"
  ))

recal_resid_df <- map2_dfr(
  recal_sim_resids,
  paste0(seq_along(recal_sim_resids)),
  ~ tibble(
    model = .y,
    site = seq_along(residuals(.x)),
    residual = residuals(.x)
  )
) %>% 
  mutate(model = case_when(
    model == "1" ~ "null_sin",
    model == "2" ~ "m1_sin",
    model == "3" ~ "m2_sin",
    model == "4" ~ "m3_sin",
    model == "5" ~ "m4_sin"
  ))

ggplot(resid_df, aes(index, residual, color = model)) +
  geom_point() +
  labs(title = "sinuosity residuals")

ggplot(recal_resid_df, aes(site, residual, color = model)) +
  geom_point() +
  labs(title = "sinuosity aggregated residuals")
```

```{r residual plots for complexity}
sim_resids <- sim_resids_comp

# recalc residuals aggregated by site
recal_sim_resids <- purrr::map(sim_resids, 
                               ~DHARMa::recalculateResiduals(.x, group = ~site),
                               .progress = FALSE)

# plotting QQ plot and predicted vs residuals aggregated by site
# makes for a low sample size of residuals 
#purrr::map(recal_sim_resids, ~plot(.x, quantreg = FALSE))

# plotting just residuals

resid_df <- map2_dfr(
  sim_resids,
  paste0(seq_along(sim_resids)),
  ~ tibble(
    model = .y,
    index = seq_along(residuals(.x)),
    residual = residuals(.x)
  )
) %>% 
  mutate(model = case_when(
    model == "1" ~ "null_sin",
    model == "2" ~ "m1_sin",
    model == "3" ~ "m2_sin"
  ))

recal_resid_df <- map2_dfr(
  recal_sim_resids,
  paste0(seq_along(recal_sim_resids)),
  ~ tibble(
    model = .y,
    site = seq_along(residuals(.x)),
    residual = residuals(.x)
  )
) %>% 
  mutate(model = case_when(
    model == "1" ~ "null_sin",
    model == "2" ~ "m1_sin",
    model == "3" ~ "m2_sin",
    model == "4" ~ "m3_sin",
    model == "5" ~ "m4_sin"
  ))

ggplot(resid_df, aes(index, residual, color = model)) +
  geom_point() +
  labs(title = "complexity residuals")

ggplot(recal_resid_df, aes(site, residual, color = model)) +
  geom_point() +
  labs(title = "complexity aggregated residuals")
```


## stage 3: sensitivity

it could be good to ask whether our results change with the period of record we use (i.e. the imagery quality)
especially if we think the quality of wood data drastically improves in a given period

shayla can make a plot for this 