---
title: "wood_dynamism"
author: 'Noël Clark'
date: "2026-06-22"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      out.width = "100%")
```

```{r load libraries, message = FALSE}
# Adding a packageLoad function within the script "setup.R" because I didn't have all of these downloaded
# Add additional packages within the script, or just use the library function. Sorry to make this complicated
# - Shayla
source("setup.R")
```

```{r everyone load data, message = FALSE}
# this chunk should just work for everyone when you open the project. I copied the data into a subfolder called data within this project, but in case it doesn't work I left Noel and Kelley's load chunks below

dynamism_data <- read_csv("data/site_metrics_07022026.csv")
data_byyr <- read_csv("data/SelectMetricsByYear_07022026.csv") 
all_data_byyr <- read.csv("data/ALL_metrics_07022026.csv") 
```

```{r kelley load data, message = FALSE}
setwd("~/Library/CloudStorage/OneDrive-TheUniversityofMontana/Wood+/Wood+") # Kelley's working directory

dynamism_data <- read_csv("site_metrics_07022026.csv") # Kelley's path
data_byyr <- read_csv("SelectMetricsByYear_07022026.csv") # Kelley's path
all_data_byyr <- read.csv("ALL_metrics_07022026.csv") # Kelley's path
```

```{r noel load data, message = FALSE}
## change directory if needed ##
data_dir <- "C:/Users/nc274936/OneDrive - The University of Montana/wood_dynamism/data"

dynamism_data <- read_csv(file.path(data_dir, "Metrics/site_metrics_07022026.csv"))
data_byyr <- read_csv(file.path(data_dir, "/Metrics/SelectMetricsByYear_07022026.csv"))
all_data_byyr <- read.csv(file.path(data_dir, "/Metrics/ALL_metrics_07022026.csv"))
```


```{r clean loaded data}
## this should work after you run the right version of data loading above
dynamism_data <- dynamism_data[1:13,] %>% # for some reason we have duplicate rows in the csv
  bind_rows( # also want to add back the mod rock_WA site
    dynamism_data %>% slice(6) %>% 
      mutate(across(
        c(site_reach, site),
        ~ str_replace_all(.x, "_1", "_2")
        ))
    ) 

data_byyr <- data_byyr %>% 
  rename(reach = site,
         site = river) %>% 
  mutate(site_reach = paste0(site,reach)) %>% 
  mutate(ratio_accum_to_ch_area = accum_area/channel_area,
         piece_per_100m = (wood_pieces/channel_length)*100,
         site = factor(site),
         reach = factor(reach))

all_data_byyr <- all_data_byyr %>% 
  rename(reach = site,
         site = river) %>% 
  mutate(site_reach = paste0(site,reach))
```

# Simple Plots
 
plot settings
```{r plot settings}
wood_colors = c("LW_Channel" = "#99C5B5", "no_LW_Channel" = "#BEA57D")
```

# visualizing distribution of response metrics 

support for the response values is positive & continuous
none of the response values are normally distributed - may require transformation?

support for the predictor is between 0 and 1 (proportion)
the predictor follows a negative binomial (?) distribution

```{r simple response value plots}
data_byyr %>% 
  ggplot(aes(x=sinuosity)) +
  geom_density() 

data_byyr %>% 
  ggplot(aes(x=complexity)) +
  geom_density()

data_byyr %>% 
  ggplot(aes(x=sdi)) +
  geom_density() 

data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area)) +
  geom_density() 

data_byyr %>%  
  ggplot(aes(x=piece_per_100m)) +
  geom_density() 
```

## stage 1: wood as a response to site, reach, & time
```{r wood v time plots}
# how does standardized wood (the ratio of wood accumulation to channel area) change over time?
data_byyr %>%
  ggplot(aes(x=year, y=ratio_accum_to_ch_area, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach))

# how does the difference between standardized wood change within a site (between reaches) over time? (diff = LW - noLW)
data_byyr %>%
  group_by(site, year) %>% 
  arrange_all(by_group = "reach") %>% 
  mutate(accum_ratio_diff = lag(ratio_accum_to_ch_area, 1) - ratio_accum_to_ch_area) %>% 
  dplyr::select(site, year, accum_ratio_diff) %>% 
  drop_na(accum_ratio_diff) %>% 
  ggplot(aes(x=year, y=accum_ratio_diff, color = site)) +
  geom_point() +
  geom_line(aes(group = site))


# how does standardized wood pieces change over time?
data_byyr %>% 
  ggplot(aes(x=year, y=piece_per_100m, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach))

# how does the difference between standardized number of wood pieces change within a site (between reaches) over time? (diff = LW - noLW)
data_byyr %>%
  group_by(site, year) %>% 
  arrange_all(by_group = "reach") %>% 
  mutate(piece_100m_diff = lag(piece_per_100m, 1) - piece_per_100m) %>% 
  dplyr::select(site, year, piece_100m_diff) %>% 
  drop_na(piece_100m_diff) %>% 
  ggplot(aes(x=year, y=piece_100m_diff, color = site)) +
  geom_point() +
  geom_line(aes(group = site))
```
```{r simple wood site plots}
##### wood values by site by year #####
data_byyr %>% 
  ggplot(aes(x = year, y = ratio_accum_to_ch_area, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors) +
  theme(legend.position = "none") 

data_byyr %>% 
  ggplot(aes(x = year, y = piece_per_100m, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

##### mean wood values by site #####
dynamism_data %>% 
  ggplot(aes(x = site, y = avg_wood_pieces_per100m, fill = reach)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = wood_colors) +
  theme(legend.position = "none")

dynamism_data %>% 
  ggplot(aes(x = site, y = mean_wood_accum_per_area, fill = reach)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = wood_colors)

```


## stage 2: how do sinuosity, complexity, and sdi change as a function of standardized wood?

visualizing relationship between response v predictor (ratio_accum_to_ch_area)
within a site the relationship is linear-ish? 
```{r visualize sin comp and sdi v wood}
data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area, y=sinuosity, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site))

data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area, y=complexity, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site))

data_byyr %>% 
  ggplot(aes(x=ratio_accum_to_ch_area, y=sdi, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site))


## wood piece_per_100m
data_byyr %>% 
  ggplot(aes(x=piece_per_100m, y=sinuosity, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site))

data_byyr %>% 
  ggplot(aes(x=piece_per_100m, y=complexity, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site))

data_byyr %>% 
  ggplot(aes(x=piece_per_100m, y=sdi, color = site, shape = reach)) +
  geom_point() +
  geom_line(aes(group=site))
```

channel values by site
```{r simple channel site plots}
# channel characteristics v time by site
data_byyr %>% 
  ggplot(aes(x = year, y = channel_length, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

data_byyr %>% 
  ggplot(aes(x = year, y = channel_area, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) + 
  scale_color_manual(values = wood_colors)
```

response values by site
```{r simple response values by site}
# response metrics v time by site
data_byyr %>% 
  ggplot(aes(x = year, y = sinuosity, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

data_byyr %>% 
  ggplot(aes(x = year, y = complexity, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

data_byyr %>% 
  ggplot(aes(x = year, y = sdi, color = reach)) +
  geom_point() +
  geom_line(aes(group = site_reach)) +
  facet_wrap(~site) +
  scale_color_manual(values = wood_colors)

# all_data_byyr %>% 
#   ggplot(aes(x = year_to, y = net_change, color = reach)) +
#   geom_point() +
#   geom_line(aes(group = site_reach)) +
#   facet_wrap(~site) +
#   scale_color_manual(values = wood_colors)
# 
# all_data_byyr %>% 
#   ggplot(aes(x = year_to, y = erosion_area, color = reach)) +
#   geom_point() +
#   geom_line(aes(group = site_reach)) +
#   facet_wrap(~site) +
#   scale_color_manual(values = wood_colors)
# 
# all_data_byyr %>% 
#   ggplot(aes(x = year_to, y = deposition_area, color = reach)) +
#   geom_point() +
#   geom_line(aes(group = site_reach)) +
#   facet_wrap(~site) +
#   scale_color_manual(values = wood_colors)
```

dynamism values by site
```{r simple dynamism plots}
##### Sinuosity vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = sinuosity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "Sinuosity Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = sinuosity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "Sinuosity Dynamism vs. Avg Wood Metrics")
##### end 

##### Complexity vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = complexity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "Complexity Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = complexity_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "Complexity Dynamism vs. Avg Wood Metrics")
##### end

##### SDI vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = sdi_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "SDI Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = sdi_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "SDI Dynamism vs. Avg Wood Metrics")
##### end

##### Net change vs both mean wood metrics #####
p1 <- dynamism_data %>%
  ggplot(aes(x = avg_wood_pieces_per100m, y = net_change_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() + 
  theme(legend.position = "none") +
  labs(x = "Avg. Wood Pieces per 100m", y = "Net Change Dynamism")

p2 <- dynamism_data %>%
  ggplot(aes(x = mean_wood_accum_per_area, y = net_change_dynamism, color = site)) +
  geom_point(aes(shape = reach)) +
  geom_line(aes(group = site), color = "black", alpha = 0.5) +
  theme_bw() +
  labs(x = "Avg. Wood Acculumations per Area", y = "")

(p1 + p2) +
  plot_annotation(title = "Net Change Dynamism vs. Avg Wood Metrics")
##### end
```

```{r simple dynamism plots pt. 2}
library(grid)
library(rcartocolor)
library(dplyr)
library(patchwork)

## CLEANING UP
# Adjusting so all 3 metrics can be plotted on same graph
dynamism_cubed <- dynamism_data %>%
  pivot_longer(c(complexity_dynamism,sinuosity_dynamism, sdi_dynamism, net_change_dynamism),
               names_to = "Metric",
               values_to = "Value") %>%
  group_by(Metric) %>%
  mutate(Value_scaled = (scale(Value))) # calculated a separate z-score for each dynamism metric
  # will show dynamism values relative to each other

# Essentially, does wood accumulation change dynamism relative to it's normal range?

# Looking at colorblind friendly color schemes
#rcartocolor::display_carto_all()
#carto_pal(7, "Earth")

# Renaming labels
dynamism_cubed <- dynamism_cubed %>%
  mutate(
    Metric = recode(
      Metric,
      complexity_dynamism = "Complexity",
      sinuosity_dynamism = "Sinuosity",
      sdi_dynamism = "SDI",
      net_change_dynamism = "Net Channel Change"
    ),
    reach = recode(
      reach,
      LW_Channel = "Large Wood",
      no_LW_Channel = "Control"
    ))
    
## PLOTS

# Mean wood pieces (pooled)
ggplot(dynamism_cubed, aes(avg_wood_pieces_per100m, Value_scaled, color = Metric)) +
  geom_line(size = 1) +
  scale_color_manual(values = c(
  "Complexity" = "#B5B991",
  "Sinuosity" = "#70A494",
  "SDI" = "#DE8A5A",
  "Net Channel Change" = "#E15759"
  )) +
  labs(x = "Mean wood pieces (per 100m)",
    y = "Standardized dynamism (z-score)") +
  theme_classic()

# Mean wood accumulation (pooled)
ggplot(dynamism_cubed, aes(mean_wood_accum_per_area, Value_scaled, color = Metric)) +
  geom_line(size = 1)+
  scale_color_manual(values = c(
  "Complexity" = "#B5B991",
  "Sinuosity" = "#70A494",
  "SDI" = "#DE8A5A",
  "Net Channel Change" = "#E15759"
  )) +
  labs(x = "Mean wood accumulation (per area)",
    y = "Standardized dynamism (z-score)") +
  theme_classic()

## COMBO PLOT
# Mean wood peices across site and reach
meanwoodpieces <- ggplot(
  dynamism_cubed,
  aes(
    x = avg_wood_pieces_per100m,
    y = Value_scaled,
    color = Metric
  )
) +
  scale_color_manual(values = c(
  "Complexity" = "#B5B991",
  "Sinuosity" = "#70A494",
  "SDI" = "#DE8A5A",
  "Net Channel Change" = "#E15759"
  ))+
 geom_point(
    aes(shape = site),
    size = 3)+
  scale_shape_manual(values = c(16, 17, 15, 18, 3, 8, 4)) +
  geom_smooth(aes(group = Metric),
              method = "lm",
              se = FALSE,
              linewidth = 1.2) +
  facet_wrap(~reach) +
    facet_wrap(~reach, scales = "free_x") + # delete scale = "free_x" if you want each reach on the same scale
    labs(
    x = "Mean wood pieces (per 100m)",
    y = "Standardized dynamism (z-score)",
    color = "Dynamism Metric",
    shape = "Site")+
  theme_bw()


# Mean wood accumulation across site and reach
meanwoodaccum <- ggplot(
  dynamism_cubed,
  aes(
    x = mean_wood_accum_per_area,
    y = Value_scaled,
    color = Metric
  )
) +
  scale_color_manual(values = c(
  "Complexity" = "#B5B991",
  "Sinuosity" = "#70A494",
  "SDI" = "#DE8A5A",
  "Net Channel Change" = "#E15759"
  ))+
 geom_point(
    aes(shape = site),
    size = 3)+
  scale_shape_manual(values = c(16, 17, 15, 18, 3, 8, 4)) +
  geom_smooth(aes(group = Metric),
              method = "lm",
              se = FALSE,
              linewidth = 1.2) +
    facet_wrap(~reach, scales = "free_x") + # delete scale = "free_x" if you want on the same scale
    labs(
    x = "Mean wood accumulation (per area)",
    y = "Standardized dynamism (z-score)",
    color = "Dynamism Metric",
    shape = "Site")+
  theme_bw()+
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )


## FINAL PLOT
(meanwoodpieces + meanwoodaccum) +
  plot_layout(guides = "collect") +
  plot_annotation(title = "Dynamism vs. Average Wood Metrics") &
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    legend.key.size = unit(0.4, "cm"),
    legend.spacing.x = unit(0.2, "cm"),
    legend.text = element_text(size = 9),
    legend.box.spacing = unit(0.1, "cm"),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  ) &
  guides(
    color = guide_legend(
      title = "Dynamism Metric",
      ncol = 1
    ),
    shape = guide_legend(
      title = "Site",
      ncol = 1,
      override.aes = list(size = 3)
    )
  )



## NOW, ON THE SAME X SCALE

meanwoodpieces_samex <- ggplot(
  dynamism_cubed,
  aes(
    x = avg_wood_pieces_per100m,
    y = Value_scaled,
    color = Metric
  )
) +
  scale_color_manual(values = c(
  "Complexity" = "#B5B991",
  "Sinuosity" = "#70A494",
  "SDI" = "#DE8A5A",
  "Net Channel Change" = "#E15759"
  ))+
 geom_point(
    aes(shape = site),
    size = 3)+
  scale_shape_manual(values = c(16, 17, 15, 18, 3, 8, 4)) +
  geom_smooth(aes(group = Metric),
              method = "lm",
              se = FALSE,
              linewidth = 1.2) +
  facet_wrap(~reach) +
    facet_wrap(~reach) + # delete scale = "free_x" if you want each reach on the same scale
    labs(
    x = "Mean wood pieces (per 100m)",
    y = "Standardized dynamism (z-score)",
    color = "Dynamism Metric",
    shape = "Site")+
  theme_bw()


meanwoodaccum_samex <- ggplot(
  dynamism_cubed,
  aes(
    x = mean_wood_accum_per_area,
    y = Value_scaled,
    color = Metric
  )
) +
  scale_color_manual(values = c(
  "Complexity" = "#B5B991",
  "Sinuosity" = "#70A494",
  "SDI" = "#DE8A5A",
  "Net Channel Change" = "#E15759"
  ))+
 geom_point(
    aes(shape = site),
    size = 3)+
  scale_shape_manual(values = c(16, 17, 15, 18, 3, 8, 4)) +
  scale_x_continuous(
  labels = scales::label_number(accuracy = 0.1) # putting this panel to the same decimal point as mean wood peices
)+
  geom_smooth(aes(group = Metric),
              method = "lm",
              se = FALSE,
              linewidth = 1.2) +
    facet_wrap(~reach) + 
    labs(
    x = "Mean wood accumulation (per area)",
    y = "Standardized dynamism (z-score)",
    color = "Dynamism Metric",
    shape = "Site")+
  theme_bw()+
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

(meanwoodpieces_samex + meanwoodaccum_samex) +
  plot_layout(guides = "collect") +
  plot_annotation(title = "Dynamism vs. Average Wood Metrics") &
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    legend.key.size = unit(0.4, "cm"),
    legend.spacing.x = unit(0.2, "cm"),
    legend.text = element_text(size = 9),
    legend.box.spacing = unit(0.1, "cm"),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  ) &
  guides(
    color = guide_legend(
      title = "Dynamism Metric",
      ncol = 1
    ),
    shape = guide_legend(
      title = "Site",
      ncol = 1,
      override.aes = list(size = 3)
    )
  )

```


# correlation plots and test 

ggpairs uses the pearson correlation test by default
```{r correlation plot test}
# ggpairs uses pearson cor test by default 
data_byyr %>%
  dplyr::select(sinuosity, complexity, sdi, channel_length, channel_area, accum_area, ratio_accum_to_ch_area, wood_pieces, piece_per_100m) %>% 
  ggpairs(progress = F)

dynamism_data %>% 
  dplyr::select(sinuosity_dynamism, complexity_dynamism, sdi_dynamism, net_change_dynamism, avg_wood_pieces_per100m, mean_wood_accum_per_area) %>% 
  ggpairs(progress = F)
```

# Models

"random effects should reflect the design, but complexity should be justified by the amount of information available"
experimental design is such that reaches are nested within sites to account for multiple observations (across years) at reach_site combinations 

```{r create model df}
mod_data_byyr <- data_byyr %>% 
  mutate(ratio_accum_scaled = scale(ratio_accum_to_ch_area, 
                                    center = TRUE, # centers ratio values around mean ratio which is then set to 0 
                                    scale = TRUE), # scale changes a value to the # of sd from the mean value
         piece_100m_scaled = scale(piece_per_100m, 
                                   center = TRUE, 
                                   scale = TRUE),
         year_c = scale(year, # centers years around a mean (set to 0) value for year, makes for a more interpretable effect
                        center = TRUE, scale = FALSE)) %>% 
  filter(site_reach != "Rock_WA_2no_LW_Channel") %>% 
  mutate(site = str_replace_all(site, "_[:digit:]$", ""))
```

## modeling notes
- AICc is preferred for small sample sizes which is when the number of data points / number of parameters < 40 (n/k < 40). we have 126 observations, which for the null model means we're not in the small sample size world (k=3), but for all other models we are because we have to estimate at least 4 parameters. it's good to use the same criteria across models so we'll use AICc for all. the aictab function in the AICcmodavg does this for us.  
- all of our response variables are not normally distributed so we're using a generalized linear mixed model with a Gamma family and log link unless otherwise stated

## models for sinuosity
```{r specify sinuosity candidate model}
## null model: random effect of site only reflects hierarchical design 
# this will estimate a random intercept of sinuosity for every site (i.e. mean sinuosity per site)
null_sin <- glmer(sinuosity ~ (1|site), mod_data_byyr, family = Gamma(link = "log"))

## models with scaled ratio of wood accumulation
m1_sin <- glmer(sinuosity ~ ratio_accum_scaled + (1|site), mod_data_byyr, family = Gamma(link = "log"))
m2_sin <- glmer(sinuosity ~ piece_100m_scaled + (1|site), mod_data_byyr, family = Gamma(link = "log"))

m3_sin <- glmer(sinuosity ~ ratio_accum_scaled + piece_100m_scaled + (1|site), mod_data_byyr, family = Gamma(link = "log"))
m4_sin <- glmer(sinuosity ~ ratio_accum_scaled * piece_100m_scaled + (1|site), mod_data_byyr, family = Gamma(link = "log"))

## models with more complicated random effects structure
# these calculate both a random intercept and a random slope within a site
# which on second thought I don't think we actually want 
# because sites are considered a "random" sample of the global population of sites 
# and we're interested in the effect of wood across this global population of sites, not within sites?
m5_sin <- glmer(sinuosity ~ ratio_accum_scaled + (1 + ratio_accum_scaled|site), mod_data_byyr, family = Gamma(link = "log"))
m6_sin <- glmer(sinuosity ~ piece_100m_scaled + (1 + piece_100m_scaled|site), mod_data_byyr, family = Gamma(link = "log"))

m7_sin <- glmer(sinuosity ~ ratio_accum_scaled + piece_100m_scaled + (1 + ratio_accum_scaled|site), mod_data_byyr, family = Gamma(link = "log"))
m8_sin <- glmer(sinuosity ~ ratio_accum_scaled * piece_100m_scaled + (1 + ratio_accum_scaled|site), mod_data_byyr, family = Gamma(link = "log"))
m9_sin <- glmer(sinuosity ~ ratio_accum_scaled + piece_100m_scaled + (1 + piece_100m_scaled|site), mod_data_byyr, family = Gamma(link = "log"))
m10_sin <- glmer(sinuosity ~ ratio_accum_scaled * piece_100m_scaled + (1 + piece_100m_scaled|site), mod_data_byyr, family = Gamma(link = "log"))

# full candidate set of models
# currently does not include the more complicated random effect structure
cand.set <- list(null_sin, m1_sin, m2_sin, m3_sin, m4_sin)

```

```{r DHARMa checks for candidate sinuosity models}
## model diagnostics with DHARMa & performance packages
# Plot simulated residuals: (QQ-plot for distribution, Residual vs. Predicted for variance)
sim_resids <- purrr::map(cand.set, 
                         ~DHARMa::simulateResiduals(fittedModel = .x), 
                         .progress = FALSE)

# plotting QQ plot and predicted vs residuals
# obviously, the predicted vs residuals plots look a lil weird
purrr::map(sim_resids, 
           ~plot(.x, quantreg = FALSE),
           .progress = FALSE)

# recalc residuals aggregated by site
recal_sim_resids <- purrr::map(sim_resids, 
                               ~DHARMa::recalculateResiduals(.x, group = ~site),
                               .progress = FALSE)

# plotting QQ plot and predicted vs residuals aggregated by site
# makes for a low sample size of residuals 
#purrr::map(recal_sim_resids, ~plot(.x, quantreg = FALSE))

# plotting just residuals, 

resid_df <- map2_dfr(
  sim_resids,
  paste0(seq_along(sim_resids)),
  ~ tibble(
    model = .y,
    index = seq_along(residuals(.x)),
    residual = residuals(.x)
  )
) %>% 
  mutate(model = case_when(
    model == "1" ~ "null_sin",
    model == "2" ~ "m1_sin",
    model == "3" ~ "m2_sin",
    model == "4" ~ "m3_sin",
    model == "5" ~ "m4_sin"
  ))

recal_resid_df <- map2_dfr(
  recal_sim_resids,
  paste0(seq_along(recal_sim_resids)),
  ~ tibble(
    model = .y,
    site = seq_along(residuals(.x)),
    residual = residuals(.x)
  )
) %>% 
  mutate(model = case_when(
    model == "1" ~ "null_sin",
    model == "2" ~ "m1_sin",
    model == "3" ~ "m2_sin",
    model == "4" ~ "m3_sin",
    model == "5" ~ "m4_sin"
  ))

ggplot(resid_df, aes(index, residual, color = model)) +
  geom_point() 

ggplot(recal_resid_df, aes(site, residual, color = model)) +
  geom_point() 
```

```{r performance checks for candidate sinuosity models}
# plots various model assumptions (normality of residuals, normality of random effects, linear relationship, homogeneity of variance, multicollinearity).
# some of the models have flat-ish reference lines for homogeneity of variance
performance::check_model(null_sin)
performance::check_model(m1_sin)
performance::check_model(m2_sin)
performance::check_model(m3_sin)
performance::check_model(m4_sin)

# acting unstable for some reason, think the plots are too mem heavy for map
# purrr::map(cand.set, ~performance::check_model(.x))
```

```{r sinuosity model selection}
AICcmodavg::aictab(cand.set, modnames = c("null_sim","m1_sin","m2_sin","m3_sin", "m4_sin"))

tab_model(m1_sin)
```

```{r plot top sinuosity model predictions}
# ## plot the null model
# # create plotting df with our sampled and estimated site means
# plot_null_sin <- mod_data_byyr %>% 
#   group_by(site) %>% 
#   summarize(sampled = mean(sinuosity)) %>% 
#   mutate(estimated = exp(coef(null_sin)$site$"(Intercept)"))
# 
# # solid lines are sampled means, dashed lines are estimated means
# ggplot(mod_data_byyr) +
#   geom_point(aes(x = ratio_accum_scaled, y = sinuosity, color = site)) +
#   geom_hline(aes(yintercept = sampled, color = site), plot_null_sin) +
#   geom_hline(aes(yintercept = estimated, color = site), plot_null_sin, linetype = "dashed") 

# plot model predictions example
plot(
  ggpredict(
    m1_sin, terms = c("ratio_accum_scaled", "site"), type = "random", bias_correction = TRUE
    )
  ) +
  geom_point(
    data = mod_data_byyr,
    aes(x = ratio_accum_scaled, y = sinuosity, color = site),
    alpha = 0.5, inherit.aes = FALSE
  ) +
  labs(title = "predicted & observed values of sinuosity")

```

## models 
```{r specify all models together possibly}
responses <- c("sinuosity", "complexity", "sdi")

fit_candidates <- function(response, data) {
  list(
    null = glmer(reformulate("(1|site)", response), 
                 data, family = Gamma("log")),
    m1   = glmer(reformulate("ratio_accum_scaled + (1|site)", response), 
                 data, family = Gamma("log")),
    m2   = glmer(reformulate("piece_100m_scaled + (1|site)", response), 
                 data, family = Gamma("log")),
    m3   = glmer(reformulate("ratio_accum_scaled + piece_100m_scaled + (1|site)", response), 
                 data, family = Gamma("log")),
    m4   = glmer(reformulate("ratio_accum_scaled * piece_100m_scaled + (1|site)", response), 
                 data, family = Gamma("log"))
  )
}

models <- map(set_names(responses), fit_candidates, data = mod_data_byyr)

AICcmodavg::aictab(models$sinuosity)
AICcmodavg::aictab(models$complexity)
AICcmodavg::aictab(models$sdi)
```



## stage 3: sensitivity

it could be good to ask whether our results change with the period of record we use (i.e. the imagery quality)
especially if we think the quality of wood data drastically improves in a given period